Neural Models for Target-Based Computer-Assisted Musical Orchestration: A Preliminary Study

### 2020-08-18 paper/aimusic2020/paper.tex
remove old bibliography
Cutting out some things

\pagestyle{headings}

@@

The codebase for this paper and some audio examples can be found at: \url{https://github.com/dzluke/DeepOrchestration}.

@@

\begin{figure}
	\centering
	\includegraphics[scale=0.22]{../latex/figs/CNN_LSTM_diagram.png}
	\caption{Diagram of the CNN with LSTM architecture. \label{cnn_lstm}}
\end{figure}

@@

For a given number of instruments $N$, each input to our models is a combination of $N$ TinySOL samples chosen among an orchestra of 10 instruments. The data is generated in the following way:

\begin{itemize}
\item $N$ random TinySOL samples are selected, leading to a variety of instruments, pitches, and dynamics; we did not allow the same instrument to be chosen more than three times in order to ensure variety in the mixtures;
\item the chosen samples are combined to be played simultaneously and normalized by the number of instruments;
\item the Mel frequency cepstral coefficients (MFCCs) and the Mel spectrogram of the mixture are computed to be used as the input features for the baseline classifiers and the neural classifiers respectively.
\end{itemize}

The resulting combination has a sample rate of $44100$Hz and is padded or trimmed to be exactly 4 seconds long.

@@

The complete problem of classification of the pairs instrument-pitch has 424 total classes. This problem is very hard for parametric classifiers, so we initially switched to the simpler problem of classifying only the instruments and not the pitch. This had the benefit of both reducing the number of classes and increasing the number of samples per class. We found that SVM was able to very accurately identify the instrument given an input that had only one instrument present; for this case the accuracy was 99.8\%. However, as soon as multiple instruments were present in the input, the accuracy dropped significantly. With two instruments, accuracy was 55.4\%, with three it was 19.6\%. KNN performed significantly worse than SVM, so we did not attempt any further testing with it.

To better approximate our original problem of identifying instrument and pitch, however, we performed experiments in which two instruments were selected and for input data that contained samples from one of those two instruments, both the instrument and pitch \emph{class} of the sample would be classified. The pitch class is the note name without the octave, e.g. C, D\#, G, etc. The input was a combination of two instruments drawn from a possible twelve instruments, and the classifier attempted to identify which instruments were present and for the specified instruments, say Violin and Viola, which pitch classes were present. If another instrument was present in the input combination that was not Violin or Viola, the classifier would simply identify that an instrument that was not one of the two was present. The accuracies from this experiment are outlined in Table \ref{tab:baselines}. Since RF performed worse than SVM in every experiment, we stopped testing with it and used SVM from that point on.

@@

\begin{table}
  \centering
  \caption{Comparison of accuracies between SVM and RF. Each data point is a combination of two TinySOL samples, where at least one of the samples is from one of the two instruments specified for that experiment. For the samples drawn from one of the two instruments, the pitch class of that sample is identified. For a sample not from one of the two instruments, the classifier simply attempted to identify that a sample from one of the non-specified instruments was present. In this setting, there are 25 classes: 24 from the 12 pitch classes from 2 instruments, and 1 class that specified whether an instrument that is not one of the two specified is present. \label{tab:baselines}}
    \begin{tabular}{|c|c|c|c|} 
    	  \hline
      \textbf{Instr. 1} & \textbf{Instr. 2} & \textbf{SVM Acc.} & \textbf{RF Acc.}\\
      \hline
      Violin & Flute & 38.8\% & 9.8\% \\
      \hline
      Violin & Trumpet & 33.8\% & 9.1\% \\
      \hline
      Violin & Cello & 34.8\% & 6.3\% \\
      \hline
      Cello & Viola & 32.1\% & 5.8\% \\
      \hline
      Oboe & French Horn & \textbf{39.9\%} & \textbf{17.5\%} \\
      \hline
    \end{tabular}
\end{table}

@@

Then, for both CNN with LSTM and ResNet, we computed the maximum accuracy over the epochs. 
The CNN was trained on 200,000 generated samples over 50 epochs, and ResNet was 	trained on 400,000 generated samples over 20 epochs. Fig. \ref{cnn_vs_resnet} shows the best overall test accuracy achieved by both models across the number of samples $N$.\\

@@

\begin{table*}
  \centering
     \caption{Quantitative comparison of orchestrations as ratios to Orchidea. Eqn. \ref{distance} was used with $\lambda_1 = 0.5$ and $\lambda_2 = 10$ to compute distances between orchestrations and targets. What is shown is the ratio between the distance calculated and Orchidea's distance for the same target. A value less than 1 means that the model performed worse (i.e. had a larger distance), and a value greater than 1 means the model performed better than Orchidea. The last column shows the ratio of the average distances for the model across all targets.}\label{orch_eval}
    \begin{tabular}{|c|c|c|c|c|c|c|c|c|c}
     \hline
      \textbf{Model} & \textbf{Ob + Bn} & \textbf{Bn} & \textbf{Bass cl.} & \textbf{Bell 1} & \textbf{Bell 2} & \textbf{Multiph. 1} & \textbf{Car horn} & \textbf{Boat} $\ldots$ \\
      \hline
      CNN with LSTM & 0.17 & 0.28 & 0.70 & 0.55 & 0.26 & \textbf{1.10} & 0.68 & \textbf{1.12} $\ldots$  \\
      \hline
      ResNet & 0.34 & 0.50 & 0.48 & 0.59 & 0.45 & 0.90 & 0.49 & \textbf{1.16} $\ldots$ \\
      \hline
      \hline
      \textbf{$\ldots$} & \textbf{Wind harp} & \textbf{Chord 1} & \textbf{Multiph. 2} & \textbf{Chord 2} & \textbf{Gong} & \textbf{Scream} & \textbf{Brass} & \textbf{Average}  \\
      \hline
      $\ldots$ & 0.55 & 0.79 & 0.70 & 0.57 & 0.73 & \textbf{1.14} & 0.79 & 0.71\\
      \hline
      $\ldots$  & 0.61 & 0.86 & 0.51 & 0.37 & 0.71 & \textbf{1.03} & \textbf{1.05} & 0.66 \\
      \hline
    \end{tabular}

@@

Orchidea implements many advanced features that are not supported by our models. For example, it is able to apply symbolic constraints to the search, hence allowing only specific instrumental combinations or playing styles. It is also able to reduce the search space by applying harmonic analysis on the target sound.
The dominant harmonic partials of the target are identified, and the search space is limited to only include samples of those pitches. For example, if the target is a recording of an instrument playing a C4, then the partials identified may be C4, C5, G5, and E6. The model would then only consider samples of these pitches to be used in the solution. This leads to a solution whose harmonics are much closer to the target, which is an important part of aural similarity.
Orchidea's solutions, moreover, can use any number of instruments included in the orchestra specified by the user, thus having variable-sized orchestrations from a single instrument to the whole set (this property is called \emph{sparsity} of the solution).

@@

Qualitative evaluation was done through an acoustic inspection of the solution, paying close attention to timbre and pitch. For targets that had harmonic content, it was noted if the partials present in the target were also represented in the orchestrated solution. For example, one of the samples of a bell had partials that loosely represented a C minor chord, so we checked if the orchestration contained the notes of the chord. If a target had specific notes, we identified if the note or its partials were present. For example there is target of an Oboe playing an A4 and a Bassoon playing a C\#3, the solution from ResNet contained the partials of the Bassoon's note: C\#3, G\#4, C\#5, and E6 were all included on various instruments in the solution.

@@

In the first summation, $\lambda_1$ is multiplied by all the distances calculated when there was more energy in the target than the solution, since $\delta_{k1} = 1$ only when $x_k \ge \tilde{x}_k$. Similarly in the second summation, $\lambda_2$ is multiplied by all the distances when the solution provided more energy to a frequency than the target. Therefore, the relation between $\lambda_1$ and $\lambda_2$ determines whether a solution is penalized more for undershooting or overshooting the target.

@@

\begin{equation}\label{distance}
d(x, \tilde{x}) =\lambda_1 \sum_k \delta_{k1}(x_k - \tilde{x}_k) + \lambda_2 \sum_k \delta_{k2}|x_k - \tilde{x	}_k| \\
\end{equation}
where 
$
\delta_{k1} = 
\begin{cases}
1, \text{if   } x_k \ge \tilde{x}_k \\
0, \text{otherwise}
\end{cases} 
$, $
\delta_{k2} = 
\begin{cases}
1, \text{if   } x_k < \tilde{x}_k \\
0, \text{otherwise}
\end{cases}.
$\\

While our model is not able to outperform Orchidea, it shows consistent results. We find that the CNN and ResNet give similar accuracies during training, but perform differently when tasked with orchestrating targets. Overall, CNN seems to better emulate the timbre in its orchestrations, where ResNet is better for recreating the harmonic content of the target.

@@

\iffalse
\begin{figure}
\includegraphics[scale=0.6]{figs/CNN10.png}
\end{figure}
\fi

\bibliographystyle{apacite}
\bibliography{paper}
% For bibtex users:
%\bibliography{paper}
%\bibliographystyle{apalike}
%\begin{thebibliography}{99}
%
%\bibitem{FerVic2013} Jose David Fern\'andez and Francisco Vico. {AI Methods in Algorithmic Composition: A Comprehensive Survey}. Journal of Artificial Intelligence Research, vol.48, pp. 513--582, 2013.
%
%\bibitem{Ari2005} Christopher Ariza, {Navigating the landscape of computer aided algorithmic composition systems: A definition, seven descriptors, and a lexicon of systems and research}. In Proceedings of the International Computer Music Conference, 2005.
%
%\bibitem{Maresz2003}
%Yan Maresz, {On Computer-Assisted Orchestration}. Contemporary Music Review, vol. 32, n. 1, pg. 99-109, Routledge, 2013.
%
%\bibitem{Carpentier2010}
%Gr\'egoire Carpentier, Damien Tardieu, Jonathan Harvey, G\'erard Assayag and Emmanuel Saint-James. {Predicting Timbre Features of Instrument Sound Combinations: Application to Automatic Orchestration}, Journal of New Music Research, vol. 39, n. 1, pg. 47-61, Routlegde, 2010.
%
%
%\bibitem{Benetos07}
%Emmanouil Benetos, Margarita Kotti, and Constantine Kotropoulos.
%\newblock Large scale musical instrument identification.
%\newblock Sound and Music Computing Conference, Greece, 2007.
%
%\bibitem{Bhardwaj17}
%Siddharth Bhardwaj.
%\newblock Audio data augmentation with respect to musical instrument
%  recognition.
%\newblock Master's thesis, 2017.
%
%\bibitem{Bian19}
%Wenhao Bian, Jie Wang, Bojin Zhuang, Shaojun~Wang Jiankui~Yang, and Jing Xiao.
%\newblock Audio-based music classification with densenet and data augmentation.
%\newblock arXiv preprint, https://arxiv.org/abs/1906.11620v1, 2019.
%
%\bibitem{Cella2020}
%Carmine-Emanuele Cella.
%\newblock Orchidea: a comprehensive framework for target-based assisted orchestration.
%\newblock In preparation; available on request from the author, 2020.
%
%\bibitem{Cella2020b}
%Carmine-Emanuele Cella, Daniele Ghisi, Vincent Lostanlen, Fabi\'en Levy, Yan Maresz and Joshua Fineberg.
%\newblock OrchideaSOL: a dataset of extended instrumental techniques for computer-aided orchestration.
%\newblock ICMC (under review), 2020.
%
%\bibitem{Chetry05}
%Nicolas Chetry, Mike Davies, and Mark Sandler.
%\newblock Musical instrument identification using LSF and K-means.
%\newblock Audio Engineering Society Convention 118, 2005.
%
%\bibitem{He15}
%Kaiming He, Xiangyu Zhang, Shaoqing Ren, and Jian Sun.
%\newblock Deep residual learning for image recognition.
%\newblock arXiv preprint, https://arxiv.org/abs/1512.03385, 2015.
%
%\bibitem{Hershey17}
%Shawn Hershey, Sourish Chaudhuri, Daniel P.~W. Ellis, Jort~F. Gemmeke, Aren
%  Jansen, R.~Channing Moore, Manoj Plakal, Devin Platt, Rif~A. Saurous, Bryan
%  Seybold, Malcolm Slaney, Ron~J. Weiss, and Kevin Wilson.
%\newblock CNN architectures for large-scale audio classification.
%\newblock IEEE International Conference on Acoustics, Speech and Signal Processing (ICASSP), 2017.
%
%\bibitem{Hochreiter97}
%Sepp Hochreiter and J\"{u}rgen Schmidhuber.
%\newblock Long short-term memory.
%\newblock {\em Neural Comput.}, 9(8):1735–1780, November 1997.
%
%\bibitem{Kitahara05}
%Tetsuro Kitahara, Masataka Goto, and Hiroshi~G. Okuno.
%\newblock Pitch-dependent identification of musical instrument sounds.
%\newblock Applied Intelligence 23, 267–275, Springer, 2005.
%
%\bibitem{lostanlen16}
%Vincent Lostanlen and Carmine-Emanuele Cella.
%\newblock Deep convolutional networks on the pitch spiral for musical
%  instrument recognition. ISMIR, 2016.
%
%\bibitem{mcfee15}
%Brian McFee, Colin Raffel, Dawen Liang, Daniel~P.W. Ellis, Matt McVicar, Eric
%  Battenberg, and Oriol Nieto.
%\newblock librosa: Audio and music signal analysis in python.
%\newblock {P}roceedings of the 14th {P}ython in {S}cience {C}onference, 2015.
%
%\bibitem{scikit-learn}
%Fabian Pedregosa, Ga\"el Varoquaux, Alexandre
%Gramfort, Vincent Michel, Bertrand Thirion, Olivier
%Grisel, Mathieu Blondel, Peter Prettenhofer, Ron
%Weiss, Vincent Dubourg, Jake Vanderplas, Alexandre
%Passos, David Cournapeau, Matthieu Brucher,
%Matthieu Perrot, \'Edouard Duchesnay.
%\newblock Scikit-learn: Machine learning in {P}ython.
%\newblock {\em Journal of Machine Learning Research}, 12:2825--2830, 2011.
%
%\bibitem{Salamon17}
%Justin Salamon and Juan~Pablo Bello.
%\newblock Deep convolutional neural networks and data augmentation for
%  environmental sound classification.
%\newblock {\em IEEE Signal Processing Letters}, 24(3):279–283, Mar 2017.
%
%\bibitem{McKinney2003}
%Martin F. McKinney and Jeroen Breebaart.
%\newblock Features for Audio and Music Classification.
%\newblock ISMIR, 2003.
%
%\bibitem{Cella18}
%Carmine-Emanuele Cella and Philippe Esling, {Open-source modular toolbox for computer-aided orchestration}, Proc. of the Timbre conference, Montreal, Canada, 2018.
%
%\bibitem{Gillick19}
%Jon Gillick, Carmine-Emanuele Cella and David Bamman. {Estimating Unobserved Audio Features for Target-Based Orchestration}, ISMIR, 2019.
%
%\end{thebibliography}

>>>>

\renewcommand{\baselinestretch}{0.96}
\pagestyle{headings}

@@

The codebase for this paper can be found at: \url{https://github.com/dzluke/DeepOrchestration}.

@@

%\begin{figure}
%	\centering
%	\includegraphics[scale=0.22]{../latex/figs/CNN_LSTM_diagram.png}
%	\caption{Diagram of the CNN with LSTM architecture. \label{cnn_lstm}}
%\end{figure}

@@

For a given number of instruments $N$, each input to our models is a combination of $N$ TinySOL samples chosen among an orchestra of 10 instruments. The data is generated in by selecting $N$ random TinySOL samples, leading to a variety of instruments, pitches, and dynamics; we did not allow the same instrument to be chosen more than three times in order to ensure variety in the mixtures. Then the chosen samples are combined to be played simultaneously and normalized by the number of instruments and the Mel frequency cepstral coefficients (MFCCs) and the Mel spectrogram of the mixture are computed to be used as the input features for the baseline classifiers and the neural classifiers respectively. The resulting combination has a sample rate of $44100$Hz and is padded or trimmed to be exactly 4 seconds long.

@@

We started with a simplification of our problem. We performed experiments in which two instruments were selected, and the classifier attempted to identify instrument and pitch \emph{class} of the samples. The pitch class is the note name without the octave, e.g. C, D\#, G, etc. The best accuracies from these experiments were from a combination of Oboe and French Horn, in which SVM had an accuracy of 39.9\% and RF had an accuracy of 17.5\%. Since RF performed worse than SVM in every experiment, we stopped testing with it and used SVM from that point on.

@@

%\begin{table}
%  \centering
%  \caption{Comparison of accuracies between SVM and RF. Each data point is a combination of two TinySOL samples, where at least one of the samples is from one of the two instruments specified for that experiment. For the samples drawn from one of the two instruments, the pitch class of that sample is identified. For a sample not from one of the two instruments, the classifier simply attempted to identify that a sample from one of the non-specified instruments was present. In this setting, there are 25 classes: 24 from the 12 pitch classes from 2 instruments, and 1 class that specified whether an instrument that is not one of the two specified is present. \label{tab:baselines}}
%    \begin{tabular}{|c|c|c|c|} 
%    	  \hline
%      \textbf{Instr. 1} & \textbf{Instr. 2} & \textbf{SVM Acc.} & \textbf{RF Acc.}\\
%      \hline
%      Violin & Flute & 38.8\% & 9.8\% \\
%      \hline
%      Violin & Trumpet & 33.8\% & 9.1\% \\
%      \hline
%      Violin & Cello & 34.8\% & 6.3\% \\
%      \hline
%      Cello & Viola & 32.1\% & 5.8\% \\
%      \hline
%      Oboe & French Horn & \textbf{39.9\%} & \textbf{17.5\%} \\
%      \hline
%    \end{tabular}
%\end{table}

@@

Then, for both CNN with LSTM and ResNet, we computed the maximum accuracy over the epochs. 
The CNN was trained on 200,000 generated samples over 50 epochs, and ResNet was 	trained on 400,000 generated samples over 20 epochs. Fig. \ref{cnn_vs_resnet} shows the best overall test accuracy achieved by both models across the number of samples $N$.

@@

\begin{table*}
  \centering
     \caption{Quantitative comparison of orchestrations as ratios to Orchidea. Eqn. \ref{distance} was used with $\lambda_1 = 0.5$ and $\lambda_2 = 10$ to compute distances between orchestrations and targets. What is shown is the ratio between the distance calculated and Orchidea's distance for the same target. A value less than 1 means that the model performed worse (i.e. had a larger distance), and a value greater than 1 means the model performed better than Orchidea. The last column shows the ratio of the average distances for the model across all targets.}\label{orch_eval}
	\centering    
    \begin{tabular}{|c|c|c|c|c|c|c|c|c|c}
     \hline
      \textbf{Model} & \textbf{Ob + Bn} & \textbf{Bn} & \textbf{Bass cl.} & \textbf{Bell 1} & \textbf{Bell 2} & \textbf{Multiph. 1} & \textbf{Car horn} & \textbf{Boat} $\ldots$ \\
      \hline
      CNN with LSTM & 0.17 & 0.28 & 0.70 & 0.55 & 0.26 & \textbf{1.10} & 0.68 & \textbf{1.12} $\ldots$  \\
      \hline
      ResNet & 0.34 & 0.50 & 0.48 & 0.59 & 0.45 & 0.90 & 0.49 & \textbf{1.16} $\ldots$ \\
      \hline
      \hline
      \textbf{$\ldots$} & \textbf{Wind harp} & \textbf{Chord 1} & \textbf{Multiph. 2} & \textbf{Chord 2} & \textbf{Gong} & \textbf{Scream} & \textbf{Brass} & \textbf{Average}  \\
      \hline
      $\ldots$ & 0.55 & 0.79 & 0.70 & 0.57 & 0.73 & \textbf{1.14} & 0.79 & 0.71\\
      \hline
      $\ldots$  & 0.61 & 0.86 & 0.51 & 0.37 & 0.71 & \textbf{1.03} & \textbf{1.05} & 0.66 \\
      \hline
    \end{tabular}

@@

Orchidea implements many advanced features that are not supported by our models. For example, it is able to apply symbolic constraints to the search, hence allowing only specific instrumental combinations or playing styles. It is also able to reduce the search space by applying harmonic analysis on the target sound, and create sparse solutions that do not use every instrument specified.

@@

Qualitative evaluation was done through an acoustic inspection of the solution, paying close attention to timbre and pitch. For targets that had harmonic content, it was noted if the partials present in the target were also represented in the orchestrated solution. For example, one of the samples of a bell had partials that loosely represented a C minor chord, so we checked if the orchestration contained the notes of the chord.

@@

@@

\begin{equation}\label{distance}
d(x, \tilde{x}) =\lambda_1 \sum_k \delta_{k1}(x_k - \tilde{x}_k) + \lambda_2 \sum_k \delta_{k2}|x_k - \tilde{x	}_k| \\
\end{equation}
where $\delta_{k1} = 1 \text{  if  } x_k \ge \tilde{x}_k, 0 \text{  otherwise}$; and $\delta_{k2} = 1 \text{  if  } x_k < \tilde{x}_k, 0 \text{  otherwise}$.

While our model is not able to outperform Orchidea, it shows consistent results. We find that the CNN and ResNet give similar accuracies during training, but perform differently when tasked with orchestrating targets. Overall, CNN seems to better emulate the timbre in its orchestrations, where ResNet is better for recreating the harmonic content of the target. You can listen to the targets and orchestrated solutions from Orchidea, ResNet, and the CNN with LSTM at \url{https://dzluke.github.io/DeepOrchestration/}.

@@

\bibliographystyle{apacite}
\bibliography{paper}
